\begin{abstract}
Este trabalho apresenta o resultado do desenvolvimento de uma ferramenta que utiliza técnicas de processamento de imagens e redes neurais artificiais, com objetivo de facilitar o diagnóstico de exames de tomografia computadorizada pulmonar. Possui uma interface web a qual pode ser instalada em um servidor e utilizada de qualquer lugar. É um sistema que com a interação do usuário tem a capacidade de ir se adaptando e melhorando, pois é um sistema de recuperação de imagens baseada em conteúdo que leva em consideração o usuário. Este trabalho faz uso de redes neurais artificiais com topologia de pró-alimentação e algoritmo de retro-propagação para treinamento. Dessa forma o sistema pode aprender a similaridade entre duas imagens. A ferramenta desenvolvida nesse trabalho pode dimnuir a taxa de erros nos diagnósticos de um médico inexperiente.

\end{abstract}


\begin{englishabstract}{Application for computer aided diagnosis of computed tomography lung images}{image processing, computed tomography, artificial neural network, content based image retrieval}

This paper presents the result of the development of a tool that uses some techniques of image processing and artificial neural networks, aiming to facilitate the diagnosis of lung computed tomography examinations. It has a web interface which can be installed and used on a server from anywhere. It is a system that using the user's relevance feedback is capable of adjusting and improving itself, because it is a system of content based image retrieval. This work makes use of artificial neural networks with feedforward topology and back-propagation algorithm for training. Thus the system can learn the similarity between two images. The tool developed in this work can lower the erro rate in the diagnosis of an inexperienced doctor.
\end{englishabstract}